Delayed effects of climate on vital rates leads to demographic divergence in Amazonian forest fragments

### 2021-06-14 doc/paper.Rmd
added abstract

institute:
  - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - UF_LAS: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530  
  - BDFFP: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      linenumbers: yes
      number_sections: no
      fig_caption: yes
      reference_docx: "templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
      - --lua-filter=templates/pagebreak.lua
      - -Fpandoc-crossref
linkcolor: blue
bibliography: references.bib
csl: "templates/global-change-biology.csl" # Insert path for the bib-style
abstract: |
  Text of abstract. 34,281 year x individual measurements! 
keywords: |
  Amazon; delayed effects; demography; distributed lag non-linear models; drought; ENSO; habitat fragmentation; *Heliconia acuminata*; Heliconiaceae
highlights: |
  These are the highlights

@@

>>>>

institute:
  - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - UF_LAS: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530  
  - BDFFP: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      linenumbers: yes
      number_sections: no
      fig_caption: yes
      reference_docx: "templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
      - --lua-filter=templates/pagebreak.lua
      - -Fpandoc-crossref
linkcolor: blue
bibliography: references.bib
csl: "templates/global-change-biology.csl" # Insert path for the bib-style
abstract: |
  Deforestation is a major threat to species biodiversity in the Amazon rainforest. Deforestation results in loss of habitat, but also often leaves remaining forest habitat highly fragmented, with remnants of different sizes embedded in an often highly contrasting matrix. Local extinction of species from individual fragments is common, but the demographic mechanisms underlying these extinctions are poorly understood. It is often hypothesized that the dramatically altered environmental conditions in fragments are driving declines in reproduction, recruitment, or survivorship. In addition to deforestation, the Amazon is facing changes in precipitation events due to climate change with some regions predicted to face increasingly intense and more frequent droughts and other parts of the basin predicted to see more frequent and severe high precipitation events. These changes can have detrimental consequences for plant survival and reproduction. Whether plant populations in tropical forest fragments are particularly susceptible to extremes in precipitation remains unclear. Most studies of plants in fragments are relatively short (1–6 years), focus on a single life-history stage, and often do not compare to populations in continuous forest. Even fewer of these studies take into account the possibility of delayed effects of climate despite the prevalence of such delayed effects on plant demographic vital rates in studies that look for them. Using a decade of demographic and climate data from an experimentally fragmented landscape in the Central Amazon, we assess the effects of climate on populations of an understory herb (Heliconia acuminata, Heliconiaceae). We modeled potentially delayed and non-linear effects of climate on plant vital rates without *a priori* assumptions about climate windows using distributed lag non-linear models. 
  We found that delayed detrimental effects of extremes in precipitation on survival were more intense in forest fragments compared to continuous forest. The size transitions of plants in forest fragments was also more influenced by climate compared to continuous forest, with the direction of the effect depending on the lag time and season. Flowering probability was size dependent and much lower in forest fragments due to smaller average plant size. In continuous forest, flowering probability increased during droughts and when there was drought two dry seasons prior. The complex delayed effects of climate and habitat fragmentation on H. acuminata vital rates points to the importance of long term demography experiments in understanding the effects of anthropogenic change on plant populations.
keywords: |
  Amazon; delayed effects; demography; distributed lag non-linear models; drought; ENSO; habitat fragmentation; *Heliconia acuminata*; Heliconiaceae
highlights: |
  These are the highlights

@@

Running title: `r rmarkdown::metadata$shorttile`
